---
title: "NOAA Social Media Explorer"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

As kind of a fun thing to do I started tracking Twitter activity related to a number of NOAA accounts.  I've been storing this information in a relational database that I refresh about once a week to get new activity.  I've also created a basic R Shiny App to provide a way for others to explore these data.

I thought this might be a cool thing for the RUsers to talk about since it contains elements of three R functionalities that people might be interested in:

1. R packages for accessing public data through APIs
2. The RODBC package for working with relational databases in R
3. The use of R Shiny to allow others to explore your data

## Outline

The project follows a multi-step process to display information about NOAA-related activity on Twitter.  This document is organized to provide detail on each step:

### Step 0: Set up a database

For this project I set up two tables inside a database called ECON on our lab's Pinniger server. This initial set up was done entirely with MS SQL Server Management Studio, involved no R-ing, was really simple, and further discussion is likely of no interest to anyone in the group.  Moving on.

### Step 1: Collect and organize Twitter activity from NOAA accounts

This step uses the [rtweet](http://rtweet.info/) package to manage calls to Twitter's API.  

### Step 2: Push the Twitter data up to the database

This step uses the [RODBC Package](https://cran.r-project.org/web/packages/RODBC/index.html) to connect to my databases on Pinniger and add Twitter data to existing tables. 

### Step 3: Save the updated Twitter data to local .csv files

This step is clunky and probably unnecessary in the long run...but for now I don't really want my Shiny app connecting to SQL Server databases on Pinniger so I'm saving the data locally as well as in database tables.

### Step 4: Create an R Shiny App to display the data

Using [Shiny](https://shiny.rstudio.com/) I created an interactive explorer for the NOAA Twitter data that allows users to choose a subset of NOAA accounts and a time window and outputs:

* most retweeted hashtags over the time window
* temporal evolution of hashtag popularity
* most retweeted accounts
* evolution of followers by account over the time window
* the text of the most retweeted tweets over the chosen time window


## Step 1: Collect the tweets

### Step 1A: Set up communication with the Twitter API

In order to use the [rtweet](http://rtweet.info/) package to send requests to Twitter's API you have to register an application with Twitter.  This is not difficult to do but is a little tedious to explain.

Basically, you have to go to [Twitter's Developers page](https://apps.twitter.com) and register an app with them.  When you do that, you're app will be assigned access keys that can be used to set up the connection to Twitter's API.

[Michael Kearney](http://rtweet.info/articles/auth.html) has a pretty thorough guide on setting up the rtweet package.

```{r,message=FALSE}

library(data.table)
library(dplyr)
library(ggplot2)
library(rtweet)
library(RODBC)

#------------------------------------------------------------------
#Twitter credentials

#pw: mypassword
#username: aaronmams
#email: aaron.mams@gmail.com
#app name: mamultron
#owner: mamulsaurus
#owner id: myownerid
#------------------------------------------------------------------

# Uncomment everything below here for setup
#consumer_key <- ''
#consumer_secret <- ''
#access_token <- ''
#access_secret <- ''

#twitter_token <- create_token(app='mamultron',
#                              consumer_key=consumer_key,
#                              consumer_secret=consumer_secret)
#
```


I've commented out the setup code above because the setup only needs to be done once.  This is a fun improvement over the twittR package which required authentication for every new R session.

### Step 1B: Pull some tweets

Next, I define a list of NOAA Twitter accounts that I want to pull data for:
```{r}
users <- c('@NOAAHabitat','@NOAAFisheries','@NOAAFish_WCRO','@NOAA','@NOAAResearch','@NOAANCEIclimate',
           '@NOAAClimate','@MBNMS','@NOAA_CINMS','@Eileen_NOAAFish','@NOAAFisheriesAK','@NOAAFish_PIRO',
           '@NOAAFish_PIFSC','@NOAAFish_SERO','@CenterForBioDiv')
```


The data pull happens in two parts:

1. use the get_timeline() method to get all the tweets sent from the various NOAA accounts
2. use the search_tweets() to get tweets mentioning any of the NOAA accounts in the list above

```{r,eval=FALSE}
# get tweets from the NOAA accounts
new.tweets <- get_timeline('@NOAAHabitat',n=10,retryonratelimit=TRUE)

new.tweets <- new.tweets %>% select(created_at,user_id,screen_name,text,
                    is_quote,is_retweet,favorite_count,
                    retweet_count,hashtags,mentions_screen_name) %>%
  mutate(update_time=Sys.time())

str(new.tweets)
print.data.frame(new.tweets[1:3,])



 get tweets mentioning any NOAA account
  new.mentions <- search_tweets(
    '@NOAAHabitat', n = 10, include_rts = FALSE
  )
  new.mentions <- new.mentions %>% select(created_at,user_id,screen_name,text,
                      is_quote,is_retweet,favorite_count,
                      retweet_count,hashtags,mentions_screen_name) %>% 
    mutate(update_time=Sys.time())
  
str(new.mentions)
print.data.frame(new.mentions[1:3,])

new.tweets <- rbind(new.tweets,new.mentions)
```

For illustrative purposes look at the list of accounts from each data frame above:

```{r,eval=FALSE}
unique(new.tweets$screen_name)
unique(new.mentions$screen_name)
```

### Step 1C: Pull new user info

The user info is mostly meta information that may not need to be refreshed as frequently as the actual twitter activity.  However, there is one piece of information in the user profiles that I am interested in tracking over time: followers.  So, for now, I've set up the script to refresh the user profile info each time I update the Twitter activity data.

I get the user info using the *search_user()* function from the rtweet library.

```{r,eval=FALSE}
new.user.info <- lapply(users,function(x){
  user.info <- search_users(x) %>% 
    select(user_id,name,screen_name,location,description,followers_count,
           friends_count)
  return(user.info)
})

new.user.info <- data.frame(rbindlist(new.user.info))

new.user.info <- new.user.info %>% mutate(updated=as.POSIXct(Sys.time()))

```

Note that the user profile data is just a snap-shot in time.  I'm kind of interesting in watching how followers for each account grow over time.  So the last step in the code chunk above is to create a date variable for the data frame.

### Step 1D: Fix some stuff

This last step is not necessarily pivotal to the project it just massages the data frames into a form I find a little easier to deal with.  In the "new.tweets" data frame above the field "hashtags" and "mentions" are actually lists within a data frame.  I don't really like this so I'm going to transform those columns into character strings where multiple hashtags or mentions for a observation are separated by a ";".

```{r,eval=FALSE}
#fix hashtags
hash <- unlist(lapply(new.tweets$hashtags,function(x){paste(c(unlist(x)),collapse=";")}))

new.tweets <- new.tweets %>% mutate(hash1=hash) %>%
  select(-hashtags) %>%
  mutate(hashtags=hash1) %>%
  select(-hash1)

#fix @ mentions
mentions <- unlist(lapply(new.tweets$mentions_screen_name,function(x){paste(c(unlist(x)),collapse=";")}))
new.tweets <- new.tweets %>% mutate(mentions1=mentions) %>%
  select(-mentions_screen_name) %>%
  mutate(mentions=mentions1) %>%
  select(-mentions1)


str(new.tweets)
```

Note that the "hashtags" and "mentions" fields are now string entries rather than lists. 

## Step 2: Set up/Update/Refresh the database

Ok, so this is a little out of order relative to how things are done in the R Project I posted on GitHub but I thought it might be wise to keep the RODBC stuff in a separate section from the Twitter stuff.  

For this project there are two relatively simple tasks that leverage RODBC methods:

1. Pull data from the existing database tables
2. Push new data to those database tables

The only quasi-difficult part of using R  to interact with a database in SQL Server is setting up the database connection.  I'm not going to walk through this because I don't really understand it well enough to give good advice.  If you're in a DIY mood you might check out:

[Setting up a datasource on Windows](https://support.microsoft.com/en-us/help/965049/how-to-set-up-a-microsoft-sql-server-odbc-data-source)

[If you are on a Mac, follow Eric Anderson's instructions here](http://eriqande.github.io/2014/12/19/setting-up-rodbc.html)

Otherwise, get somebody who knows what they're doing to help you (shouldn't take them more than a few minutes)

### Step 2A: Pull existing data from database tables

Assuming you have a database connection, pulling the tables containing the social media data is pretty straightforward.  Using the RODBC package you:

* tell R how to connect and with which database 
* send an SQL query
* close the connection

In my PacFIN_Econ database on the Pinniger Server, I have a table called "tweets_content" and a table called "tweets_users". The "tweets_content" table holds all the actual tweets that I've pulled from Twitter and the "tweets_users" holds the meta information on each user present in the "tweets_content" table.  

To get the existing data from my database tables I use the *odbcConnect()* function to set up communication between my R session and my database, then I use the *sqlQuery()* function to pass a SQL query in a character string.

```{r, eval=FALSE}
channel <- odbcConnect(dsn='pinniger', uid='',pwd='')
content <- sqlQuery(channel,'select * from [PacFIN_Econ].[dbo].[tweets_content]')
users.info <- sqlQuery(channel,'select * from [PacFIN_Econ].[dbo].[tweets_users]')
close(channel)
```

Note that I'm not running the code above because I don't want to display my login credentials.  But I've exported these tables to csv files so I can show what they look like when read into R:

```{r}
content <- read.csv('data/tweet_content.csv')
users.info <- read.csv('data/tweet_users.csv')

print.data.frame(content[1:5,])
print.data.frame(users.info[1:5,])
```

### Step 2B: Update the tables

The reason I pulled down the existing data in the step above is that the Twitter accounts that I'm tracking aren't super active.  This means that if I pull down say the 200 most recent tweets for two weeks in a row for the same user I'm likely to generate a lot of duplicated data.  To avoid saving a bunch of duplicates to my database table I use the 4-step process of 

1. get the existing data from the database (done in Step 2A)
2. pull the new Twitter activity (done in Step 1)
3. join the new activity with the old stuff and remove duplicates
4. save the joined dataframe in the database 


At this point in the discussion we've got items 1 and 2 from the list above knocked out.  We just need to (i) join the new data with the old data and remove duplicates and (ii) send these joined/cleaned data back to the database


Joining the new with the old and removing duplicates is relatively simple:

```{r,eval=FALSE}

content <- rbind(content,new.tweets)
users.info <- rbind(users.info,new.user.info) 

#get rid of any duplicates
content <- content %>% group_by(user_id,screen_name,created_at,text) %>%
           filter(row_number()==1)

users.info <- users.info %>%
              group_by(user_id,screen_name,updated) %>%
              filter(row_number()==1)

#save these as back up .csv file
write.csv(content,'data/tweet_content.csv')
write.csv(users.info,'data/tweet_users.csv')

```

Sending the new tables up to the database is somewhat tricky but only because there are a lot of different ways to use RODBC to write to a database table.  At the moment I'm using the simplest way I could think of:

1. drop the existing database tables
2. save the new data frames as tables

```{r,eval=FALSE}
channel <- odbcConnect(dsn='pinniger', uid='',pwd='')
sqlQuery(channel,'USE PacFIN_Econ')
cols <- sqlColumns(channel,"tweets_content")
colTypes <- as.character(cols$TYPE_NAME)
colTypes[which(colTypes=='bit')] <- 'varchar(5)'
colTypes[which(colTypes=='varchar')] <- 'varchar(255)'
names(colTypes) <- as.character(cols$COLUMN_NAME)
sqlQuery(channel,"drop table tweets_content")
sqlSave(channel,content,tablename='tweets_content',append=T,rownames=F,varTypes=colTypes)
close(channel)


channel <- odbcConnect(dsn='pinniger', uid='',pwd='')
sqlQuery(channel,'USE PacFIN_Econ')
cols.users <- sqlColumns(channel,'tweets_users')
colTypes <- as.character(cols.users$TYPE_NAME)
colTypes[which(colTypes=='bit')] <- 'varchar(5)'
colTypes[which(colTypes=='varchar')] <- 'varchar(500)'
names(colTypes) <- as.character(cols.users$COLUMN_NAME)
sqlQuery(channel,"drop table tweets_users")
sqlSave(channel,users.info,tablename='tweets_users',append=T,rownames=F,varTypes=colTypes)
close(channel)

```

I'm pretty aware of the fact that this is a solution that is not likely to age well.  As the project grows and continues to pull tweets and add them to the database the data frames are going to start getting really big.  With really big data frames I suspect that writing a new table to the database each time I refresh the data is going to take intolerably long.

I'm pretty confident that there is some kind of *sqlUpdate()* or *sqlAddRows()* type method in RODBC that could update my database tables without completely wiping the old stuff and writing new tables...but I haven't had time to investigate this fully yet.

## Step 3: Update the Shiny directory

This step is pure personal preference and laziness.  Strictly speaking one could almost certainly use RODBC methods within Shiny to pull and display these data based on some user inputs.  I tried to do this but it was hard and slow, so I chose the following workflow:

1. Pull the most current data from the database
2. Save it to a .csv file in the Shiny Directory



Pulling the most current data from the database is just another instance of the RODBC methods discussed in Step 2:

```{r,eval=FALSE}
channel <- odbcConnect(dsn='pinniger', uid='',pwd='')
tweet_content <- sqlQuery(channel,'select * from [PacFIN_Econ].[dbo].[tweets_content]')
tweet_users <- sqlQuery(channel,'select * from [PacFIN_Econ].[dbo].[tweets_users]')
close(channel)

#write these data frames to .csv files in my Shiny directory
write.csv(tweet_content,file="shiny/tweet_content.csv")
write.csv(tweet_users,file="shiny/tweet_users.csv")
```

Writing these data frames to the local Shiny directory is again totally option.  I do it because I'm deploying my Shiny app to the web using [shinyapps.io](https://www.shinyapps.io).  This means I need a single directory with (i) my Shiny code and (ii) all the data dependencies.


## Step 4: Publish the Shiny App

### Quick Review

Here is what I have:

* A data frame with tweets posted by or mentioning a variety of NOAA Fisheries Twitter accounts
* A data frame with user profile information for each account

Here is what I want my Shiny App to do:

* allow the user to pick a date range
* allow the user to pick a subset of the available accounts
* display some visual summaries of Twitter engagement for the accounts during the chosen time window 

### The Shiny App

One of the coolest things about Shiny is that the start up cost is pretty low.  It doesn't take much at all to get a simple app up and running.  A great resource for starter code is the [R Shiny Gallery](https://shiny.rstudio.com/gallery/).  Here's my app:

```{r}

library(ggplot2)
library(dplyr)
library(shiny)
library(gridExtra)
library(ggthemes)
library(data.table)

hashtags <- read.csv('hashtags.csv')
tweets <- read.csv('tweet_content.csv')
users <- read.csv('tweet_users.csv')

# Define the Look for the Shiny App
ui <- fluidPage(
  headerPanel('NOAA Social Media Engagement'),
  sidebarPanel(
    dateRangeInput('daterange','Date Range',start='2017-10-01',end='2018-02-16'),
    selectInput('metric', 'Top retweeted hashtags', c('retweets')),
    checkboxGroupInput('account','NOAA Account',
                       c('NOAAHabitat','NOAAFish_WCRO',
                         'NOAAFisheries','NOAAResearch'),
                       selected=c('NOAAFish_WCRO'))
    
  ),
  mainPanel(
    tabsetPanel(
      tabPanel("Top Hashtags",
               
               plotOutput('plot1',width="100%"),
               plotOutput('plot2',width="100%")
      ),
      tabPanel("Top Users",
               plotOutput('plot3'),
               plotOutput('plot4')
      ),
      tabPanel("Top Tweets",tableOutput('table1'))
    )    
  )
)

# The server function is where data gets collected/analyzed and displayed
server <- function(input, output) {

# first I define a base data frame containing the tweets during the specified time window
  rt <- reactive({
        tweets %>% filter(screen_name %in% input$account) %>% 
          filter(as.Date(created_at) <= input$daterange[2] & as.Date(created_at)>=input$daterange[1])  
          
        
      })  

# Next, get the base data frame for user profile information
 user.followers <- reactive({  
    tbl_df(users) %>%
      filter(screen_name %in% input$account) %>%
      filter(as.Date(updated) >= input$daterange[1] & as.Date(updated)<=input$daterange[2]) %>%
      group_by(screen_name,updated) %>%
      filter(row_number()==1) %>%
      ungroup()
  })


#Now, parse out all the hashtags.  In the current data frame each row is a tweet and the hashtags contained in that tweet are stored as a # single character string with different hashtags separated by a ";". This created a new data frame where each row is a hashtag.    
  hash <- reactive({
    
    hash.list <- list()
        for(i in 1:nrow(rt())){
          h <- strsplit(as.character(rt()$hashtags[i]),";")
          r <- rt()$retweet_count[i]
          user <- rt()$screen_name[i]
          created <- rt()$created_at[i]
          hash.list[[i]] <- data.frame(rtw=r,hashtag=unlist(h),user=user,created=created)  
        }  
    
    tbl_df(data.frame(rbindlist(hash.list))) %>%
    mutate(hashtag=tolower(hashtag)) %>%
      group_by(hashtag) %>%
      mutate(created=as.Date(created,format='%Y-%m-%d %H:%M:%S')) %>%
      mutate(total=sum(rtw))
     
  })

# One of the outputs I want is the retweets over time for each of the top hashtags.  For that I'm going to expand the hashtag data set so # that there is 1 row per hashtag per day for each day in the selected date range.

  ts <- reactive({
    hash() %>% group_by(hashtag) %>% filter(row_number()==1) %>%
    group_by(hashtag) %>% mutate(start=input$daterange[1],end=input$daterange[2]) %>%
    rowwise() %>%
    do(data.frame(hashtag=.$hashtag, created=seq(.$start,.$end,by="1 day"))) %>%
    mutate(rtweets=0) 
    
  })

  #now we need to merge the expanded data with the sampled data and create the running sum
  ht <- reactive({
    hash() %>% group_by(hashtag,created) %>%
    summarise(rtw=sum(rtw))
  })
  
    rolling <- reactive({
    
    ts() %>% left_join(ht(),by=c('hashtag','created')) %>%
      mutate(rtw=ifelse(is.na(rtw),rtweets,rtw)) %>%
      mutate(rts=max(rtw,rtweets)) %>%
      group_by(hashtag) %>% arrange(hashtag,created) %>%
      mutate(retweets=cumsum(rts)) %>%
        filter(is.na(hashtag)==F) %>%
        mutate(total=max(retweets,na.rm=T)) %>%
        ungroup() %>%
        mutate(rank=dense_rank(-total)) %>%
        filter(rank<=5) %>%
        arrange(rank)
    
  })
    
   
#the first plot I want to display is pretty simple (total retweets per hashtag for the top 20 most retweeted hashtags during the time 
#window.  Since this is a pretty simple aggregation, I don't define a separate reactive object for it, I just include the dplyr pipe #inside the ggplot call. 

  output$plot1 <- renderPlot({
    
    ggplot(hash() %>% 
             group_by(hashtag) %>% 
             summarise(retweets=sum(rtw)) %>%
             ungroup() %>%
             arrange(-retweets) %>%
             filter(row_number() <= 20),aes(x=reorder(hashtag,retweets),y=retweets)) + 
      geom_bar(stat='identity',color='blue',alpha=0.6) + coord_flip() + 
      theme_wsj() + ggtitle('Total Hashtag Retweets')
    
    
  })
 
#this next plot is a little more intense: it's a rolling cumulative sum of retweets per hashtag over the entire time window.  I could #have done this as I did above by just including the dplyr pipe inside the ggplot call...but I went ahead and define a separate reactive #object for this...mostly for debugging purposes.

  output$plot2 <- renderPlot({
    
    ggplot(rolling(),
             aes(x=created,y=retweets,color=hashtag,shape=hashtag)) + geom_line() + geom_point() + 
      theme_wsj() +  theme(legend.text=element_text(size=12)) + ggtitle('Retweets')
    
    
  })
 
 # the next 2 plots appear on the "Users" tab.  They include summary info related to individual accounts. This one just shows the total 
 # retweets per account during the time window
 
  output$plot3 <- renderPlot({
    ggplot(rt() %>% 
      group_by(screen_name) %>%
      summarise(rt_count=sum(retweet_count,na.rm=T)) %>%
      arrange(-rt_count),aes(x=screen_name,y=rt_count)) + geom_bar(stat='identity') +
      theme_wsj() + ggtitle('Total Retweets')
    
  })
  
  # this one displays the follower count for each account over the time window.
  output$plot4 <-     renderPlot({ggplot(user.followers(),
                  aes(x=as.Date(updated),y=followers_count)) + geom_line() + geom_point() + 
      facet_wrap(~screen_name,scales='free') + 
      theme_wsj() +  theme(legend.text=element_text(size=12)) + ggtitle('Followers')
  })
  
  # finally, just for fun I set up a table tab to print out some of the most popular tweets from the chosen accounts during the chosen 
  # date range.
  
  output$table1 <- renderTable(  rt() %>%
                                   filter(is_retweet==FALSE) %>%
                                   filter(is.na(hashtags)==FALSE) %>%
                                   group_by(screen_name,created_at,text,hashtags,mentions) %>%
                                   summarise(RTs=sum(retweet_count,na.rm=T)) %>%
                                   arrange(-RTs) %>%
                                   ungroup() %>%
                                   filter(row_number() <= 20)  %>%
                                   select(screen_name,created_at,RTs,text) 
  )
  
}

shinyApp(ui = ui, server = server)
```

### Publish the App

[This article](https://shiny.rstudio.com/articles/shinyapps.html) has some pretty fool proof instruction for using the *rsconnect* package to deploy your Shiny app on the web.

It's also worth looking [at this article](http://docs.rstudio.com/shinyapps.io/getting-started.html#deploying-applications).

For me, setting up the *rsconnect* package was pretty simple.  I just:

* went to https://www.shinyapps.io/ and logged in
* found my dashboard
* used the options in the profile icon to generate a token
* copied the generated code into my R Session and ran it.



```{r,eval=FALSE}
rsconnect::setAccountInfo(name='aaronmams',
			  token='',
			  secret='<SECRET>')
```
